---
title: "COVID-19 Deaths by State"
author: "[Jim Tyhurst, Ph.D.](https://www.jimtyhurst.com/)"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    hard_line_breaks: FALSE
  html_document:
    toc: true
    code_folding: hide
numbersections: true
vignette: >
  %\VignetteIndexEntry{COVID_19_Deaths_by_State}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[Source code](./covid19_deaths_by_state.Rmd).

## Read data

```{r read_state_cases, echo = FALSE, message = FALSE}
library(covid19)
library(dplyr)
library(lubridate)
library(ggplot2)

cases <- read_state_cases_nytimes()
```

## Cases by state
Plot of cumulative number of cases by state in the United States for a few select states.

```{r cases_by_state, echo = FALSE, message = FALSE}
cases %>%
  dplyr::filter(state %in% c("California", "Colorado", "Oregon", "Washington"), date > lubridate::ymd("2020-02-20")) %>% 
  ggplot(aes(x = date, y = cases, color = state)) +
    geom_point(na.rm = TRUE) +
    geom_line(aes(y = cases), na.rm = TRUE)  + 
  theme(
    panel.grid.minor = element_line(color="grey60", size=0.5),
    panel.grid.major = element_line(color="grey40", size=0.5),
    panel.background = element_rect(fill="snow2")
  ) +
  ggtitle("Confirmed cases of COVID-19 for selected western states in the United States") +
  labs(x = "Date", y = "Cases", color = "State")
```

## Deaths by state
Plot of quantity of deaths per state in the United States for a few select states.

```{r deaths_by_state, echo = FALSE, message = FALSE}
cases %>%
  dplyr::filter(state %in% c("California", "Colorado", "Oregon", "Washington"), date > lubridate::ymd("2020-02-20")) %>% 
  ggplot(aes(x = date, y = deaths, color = state)) +
    geom_point(na.rm = TRUE) +
    geom_line(aes(y = deaths), na.rm = TRUE)  + 
  theme(
    panel.grid.minor = element_line(color="grey60", size=0.5),
    panel.grid.major = element_line(color="grey40", size=0.5),
    panel.background = element_rect(fill="snow2")
  ) +
  ggtitle("Deaths from COVID-19 for selected western states in the United States") +
  labs(x = "Date", y = "Deaths", color = "State")
```

## Mortality by state
Plot of percentage of deaths per state, calculated as (deaths / cases) * 100.

```{r mortality_by_state, echo = FALSE, message = FALSE}
latest_date <- max(cases$date)
cases %>% 
  dplyr::filter(date == latest_date) %>% 
  dplyr::mutate(pct = ((deaths / cases) * 100.0)) %>% 
  dplyr::arrange(desc(pct)) %>% 
  ggplot(mapping = aes(x = state, y = pct)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90))
```

## Distribution of mortality rates
Plot of mortality rates across states, i.e. percentage of deaths for confirmed cases, in the United States.

```{r mortality_distribution, echo = FALSE, message = FALSE}
latest_date <- max(cases$date)
cases %>% 
  dplyr::filter(date == latest_date) %>% 
  dplyr::mutate(pct = ((deaths / cases) * 100.0)) %>% 
  dplyr::mutate(pct_bin = floor(pct)) %>% 
  ggplot(aes(x = pct_bin, y = stat(count))) +
    geom_bar() +
    ggtitle("Mortality rates in states in the United States") +
    labs(x = "Percentage of deaths for confirmed cases", y = "Number of states")
```
