---
title: "COVID-19 Deaths by State"
author: "[Jim Tyhurst, Ph.D.](https://www.jimtyhurst.com/)"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    hard_line_breaks: FALSE
  html_document:
    toc: true
    code_folding: hide
numbersections: true
vignette: >
  %\VignetteIndexEntry{COVID_19_Deaths_by_State}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[Source code](./covid19_deaths_by_state.Rmd).

## Read data
Reads a NY Times [dataset](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html) of COVID-19 [cases and deaths](https://github.com/nytimes/covid-19-data/blob/master/us-states.csv) by state.

```{r read_state_cases, echo = FALSE, message = FALSE}
library(covid19)
library(dplyr)
library(lubridate)
library(glue)
library(ggplot2)

state_populations <-
  read_state_populations() %>% 
  select(state = state_name, population = population_estimate)
cases <- 
  read_state_cases_nytimes() %>% 
  dplyr::left_join(state_populations, by = "state")
states_to_compare <- c(
  "Arizona", 
  "California", 
  "Colorado", 
  "Idaho", 
  "Nevada", 
  "Oregon", 
  "Washington"
)
```

## Cases by state

### Cumulative number of cases
Plot of cumulative number of cases by state in the United States for a few select states.

```{r cases_by_state, echo = FALSE, message = FALSE}
cases %>%
  dplyr::filter(
    state %in% states_to_compare, 
    date > lubridate::ymd("2020-02-20")
  ) %>% 
  ggplot(aes(x = date, y = cases, color = state)) +
    geom_point(na.rm = TRUE) +
    geom_line(aes(y = cases), na.rm = TRUE)  + 
    theme(
      panel.grid.minor = element_line(color="grey60", size=0.5),
      panel.grid.major = element_line(color="grey40", size=0.5),
      panel.background = element_rect(fill="snow2")
    ) +
    labs(
      title = "Confirmed cases of COVID-19",
      subtitle = "for selected western states in the United States",
      x = "Date", 
      y = "Cases", 
      color = "State"
    )
```

### Cumulative cases per comparable unit of population
```{r cases_by_state_per_unit, echo = FALSE, message = FALSE}
population_unit = 100000
cases %>%
  dplyr::filter(
    state %in% states_to_compare,
    date > lubridate::ymd("2020-02-20")) %>%
  dplyr::mutate(cases_per_unit = ((cases / population) * population_unit)) %>%
  ggplot(aes(x = date, y = cases_per_unit, color = state)) +
  geom_point(na.rm = TRUE) +
  geom_line(aes(y = cases_per_unit), na.rm = TRUE)  +
  theme(
    panel.grid.minor = element_line(color = "grey60", size = 0.5),
    panel.grid.major = element_line(color = "grey40", size = 0.5),
    panel.background = element_rect(fill = "snow2")
  ) +
  labs(
    title = glue(
      "Cumulative cases per ",
      format(population_unit, scientific=FALSE, big.mark=","),
      " population from COVID-19"
    ),
    subtitle = "for selected western states in the United States",
    x = "Date", 
    y = glue(
      "Cumulative Cases per ", 
      format(population_unit, scientific=FALSE, big.mark=",")
    ), 
    color = "State"
  )
```

## Deaths by state

### Cumulative deaths by state
Plot of cumulative number of deaths per state in the United States for a few select states.

```{r deaths_by_state, echo = FALSE, message = FALSE}
cases %>%
  dplyr::filter(
    state %in% states_to_compare, 
    date > lubridate::ymd("2020-02-20")
  ) %>% 
  ggplot(aes(x = date, y = deaths, color = state)) +
    geom_point(na.rm = TRUE) +
    geom_line(aes(y = deaths), na.rm = TRUE)  + 
  theme(
    panel.grid.minor = element_line(color="grey60", size=0.5),
    panel.grid.major = element_line(color="grey40", size=0.5),
    panel.background = element_rect(fill="snow2")
  ) +
  labs(
    title = "Cumulative deaths from COVID-19",
    subtitle = "for selected western states in the United States",
    x = "Date", 
    y = "Cumulative Deaths", 
    color = "State"
  )
```

### Cumulative deaths per comparable unit of population
```{r deaths_by_state_per_unit, echo = FALSE, message = FALSE}
population_unit = 100000
cases %>%
  dplyr::filter(
    state %in% states_to_compare,
    date > lubridate::ymd("2020-02-20")) %>%
  dplyr::mutate(deaths_per_unit = ((deaths / population) * population_unit)) %>%
  ggplot(aes(x = date, y = deaths_per_unit, color = state)) +
  geom_point(na.rm = TRUE) +
  geom_line(aes(y = deaths_per_unit), na.rm = TRUE)  +
  theme(
    panel.grid.minor = element_line(color = "grey60", size = 0.5),
    panel.grid.major = element_line(color = "grey40", size = 0.5),
    panel.background = element_rect(fill = "snow2")
  ) +
  labs(
    title = glue(
      "Cumulative deaths per ",
      format(population_unit, scientific=FALSE, big.mark=","),
      " population from COVID-19"
    ),
    subtitle = "for selected western states in the United States",
    x = "Date", 
    y = glue(
      "Cumulative Deaths per ", 
      format(population_unit, scientific=FALSE, big.mark=",")
    ), 
    color = "State"
  )
```

### Daily number of deaths
Plot of number of deaths per day per state in the United States for a few select states.

```{r daily_deaths_by_state, echo = FALSE, message = FALSE}
cases %>%
  dplyr::filter(
    state %in% states_to_compare, 
    date > lubridate::ymd("2020-02-20")
  ) %>% 
  dplyr::arrange(date) %>% 
  dplyr::group_by(state) %>% 
  dplyr::mutate(daily_deaths = (deaths - dplyr::lag(deaths))) %>% 
  dplyr::filter(daily_deaths >= 0) %>% 
  ggplot(aes(x = date, y = daily_deaths, color = state)) +
    geom_point(na.rm = TRUE) +
    geom_line(aes(y = daily_deaths), na.rm = TRUE)  + 
    theme(
      panel.grid.minor = element_line(color="grey60", size=0.5),
      panel.grid.major = element_line(color="grey40", size=0.5),
      panel.background = element_rect(fill="snow2")
    ) +
    labs(
      title = "Daily deaths from COVID-19",
      subtitle = "for selected western states in the United States",
      x = "Date", 
      y = "Deaths", 
      color = "State"
    )
```


## Mortality by state
Plot of percentage of deaths per state, calculated as (deaths / cases) * 100.

```{r mortality_by_state, echo = FALSE, message = FALSE}
latest_date <- max(cases$date)
cases %>% 
  dplyr::filter(date == latest_date) %>% 
  dplyr::mutate(pct = ((deaths / cases) * 100.0)) %>% 
  dplyr::arrange(desc(pct)) %>% 
  ggplot(mapping = aes(x = state, y = pct)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "State", y = "Percentage of deaths for confirmed cases")
```

## Distribution of mortality rates
Plot of mortality rates across states, i.e. percentage of deaths for confirmed cases, in the United States.

```{r mortality_distribution, echo = FALSE, message = FALSE}
latest_date <- max(cases$date)
cases %>% 
  dplyr::filter(date == latest_date) %>% 
  dplyr::mutate(pct = ((deaths / cases) * 100.0)) %>% 
  dplyr::mutate(pct_bin = floor(pct)) %>% 
  ggplot(aes(x = pct_bin, y = stat(count))) +
    geom_bar() +
    labs(
      title = "Distribution of mortality rates",
      subtitle = "in states in the United States",
      x = "Percentage of deaths for confirmed cases", 
      y = "Number of states"
    )
```
